Interspecies comparisons of Mg/Ca ratios in limpet shells

### 2024-06-27 Manuscript.qmd
cleaned up format

---
title: "Interspecies comparisons of Mg/Ca ratios in limpet shells"
author:
  - name: Niklas Hausmann
    affiliations:
      - name: Leibniz Zentrum für Archäologie
        address: Ludwig-Lindenschmit-Forum 1
        city: Mainz
        country: Germany
        postal-code: 55116
    orcid: 0000–0002–4381–8334
    email: niklas@palaeo.eu
    attributes:
      corresponding: true
  - name: Donna Surge
    affiliations:
      - name: University of North Carolina
        address: 104 South Road, 225 Geology Building
        city: Chapel Hill, NC
        country: US
        postal-code: 27599-3315
  - name: Francisco Zangrando
    affiliations:
      - name: CONICET (Consejo Nacional de Investigaciones Científicas y Técnicas)
        address: Avenida Maipú 305
        city: Ushuaia
        country: Argentina
        postal-code: V9410BJA
  - name: Angelica Tivoli
    affiliations:
      - name: CONICET (Consejo Nacional de Investigaciones Científicas y Técnicas)
        address: Avenida Maipú 305
        city: Ushuaia
        country: Argentina
        postal-code: V9410BJA
  - name: Ivan Briz-Godino
    affiliations:
      - name: 
        address: 
        city: 
        country: Spain
        postal-code: 
abstract: |
  This study provides a short reassessment of the use of Magnesium to Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface temperatures. While *Patella vulgata* along the Spanish shoreline has since then repeatedly produced reliable correlations between sea surface temperature and Mg/Ca ratios, this is not the case for other patelloid species. *Patella vulgata* and *Nacella deaureata* have been studied using Mg/Ca with mixed or contrary results. In this study, we present elemental maps of various such species together with stable oxygen isotope values for some of the specimens. Our dataset also includes specimens that were previously unsuccessful in providing significant correlations between δ^18^O and Mg/Ca ratios. By reassessing these previous specimens and including a wider range of  modern and archaeological samples from three patelloid species (*P. vulgata*, *N. deaureata*, and *N. magellanica*) we further add to the growing set of evidence for the reliable use of Mg/Ca ratios to detect palaeotemperature change and serve as a means to determine ontogenetic age and season of capture as well as to reveal locations of interest within the growth record (i.e. annual temperature minima and maxima) for the targeted analysis using δ^18^O or clumped oxygen isotope analysis.
keywords: 
  - Sclerochronology
  - Limpets
  - Elemental Ratio
  - Mg/Ca
date: last-modified
bibliography: bibliography.bib
format: docx
#  elsevier-pdf:
#    keep-tex: true
#    journal:
#      name: Palaeo^3^
#      formatting: preprint
#      model: 3p
#      cite-style: authoryear
output:
  units: cm
---

@@

  theme_set(theme_cowplot())

@@

Limpet shells are commonly found within archaeological sites and and past shorelines due to their robust carbonate structure and the long-term use of limpets as a marine food source. They have successfully been studied in the past in the context of coastal subsistence economies, site occupation on seasonal [@Shackleton1973-ij; @Parker2018-wf; @Bosch2018-ud] or long-term scale [@Ortiz2015-mr], as well as palaeotemperature [@Fenger2007-gf; @Surge2012-ba; @Wang2012-ee; @Colonese2012-ct; @Ferguson2011-zl]. Determining past sea surface temperature (SST) change largely relies on the measurement of δ^18^O-values within the calcitic parts of the shell, but attempts have been made to also use elemental ratios, such as magnesium to calcium (Mg/Ca), to have an alternative measure, that potentially provides a more accurate SST estimate than δ^18^O-values, which are also affected by changes in salinity. In addition, the data acquisition of elemental ratios — either through laser-ablation-isotope-ratio-mass-spectrometry (LA-ICP-MS) or laser induced breakdown spectroscopy (LIBS), can be much faster and cost-effective, increasing the number of specimens that can be studied overall [@Durham2017-fh; @Hausmann2023-ih].

Problematically, links between Mg/Ca ratios and SST changes in many marine mollusc shells, including limpets, have shown to be unreliable [@Surge2008-ri; @Wanamaker2008-zl; @Schone2010-yl; @Freitas2012-tx; @Graniero2015-zv; @Poulain2015-dg; @Vihtakari2017-wd]. This is particularly the case where there is little available additional information on metabolic processes, organic components of the shell matrix, intra-increment and intra-shell variability, and growth rates, which can independently and unpredictably affect Mg/Ca ratios and confound their interpretation as temperature proxy. Confusingly, multiple different temperature equations have been found for the same species (see [@Freitas2012-tx; @Vihtakari2017-wd] and references therein). Coeval specimens sharing one locality can also show differences in their relation to SST [@Hausmann2019-fi]. Where the use of Mg/Ca as a palaeotemperature proxy was successful, anomalous patterns in some specimens had still to be filtered out by hand, reducing the overall robustness of the results of those successful studies [@Ferguson2011-zl].

@@

By relying and adding onto published datasets, we were able to simultaneously avoid costs for new high-resolution δ^18^O-data, to use real-world examples, and to also provide pilot-data for areas of existing research interest. Generally establishing the usefulness and reliability of Mg/Ca as SST proxy in limpets should help to provide a platform for future research and an important stepping stone to better understand elemental ratios in other marine mollusc shells.

@@

There are two main locations that we sourced specimens from: the Beagle Channel in Tierra del Fuego (Argentina) and the island of West Ray in Orkney (UK). Four specimens were collected from the Beagle Channel in Cambaceres Bay in October 2016. The area receives around 570 mm precipitation but the channel waters are predominantly influenced by marine currents and in the period of October 2015 to October 2016 stayed around 30.7 ± 0.7 psu [@Nicastro2020-ih]. SSTs range from `r CamBayMin`ºC in August to `r CamBayMax`ºC in February.

@@

```{r SST_plot}
#| fig-cap: SST values averaged from  years. SST values were sourced from @Good2020-nl
#| fig-height: 5
#| fig-width: 5
#| label: fig-SSTs

# Plot the data
SST_df %>% 
    group_by(dataset, month) %>%
  summarise(sd_sst = sd(mean_sst, na.rm = TRUE),
            mean_sst = mean(mean_sst, na.rm = TRUE)) %>%
ggplot(aes(x = month, y = mean_sst, colour = dataset,group=dataset)) +
  geom_ribbon(aes(ymin = mean_sst - sd_sst, ymax = mean_sst + sd_sst,fill=dataset), alpha = 0.2,col="white") +
  geom_path() +
  scale_colour_manual(values = c( "#007f7f", "#EC7063")) +
scale_fill_manual(values = c( "#007f7f", "#EC7063"))+
  labs(title = "Monthly Sea Surface Temperature",
       x = "Month",
       y = "Mean Analysed SST (°C)") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1),
    legend.position = c(0.1, 0.9))+  
  scale_y_continuous(breaks = c(5, 10, 15),limits = c(3,15)) +
  labs(colour="",fill="")

@@

High-quality oxygen isotope values (δ^18^O) were acquired from previous publications on *P. vulgata* [@Surge2012-ba; @Graniero2017-io], *N. deaureata* and *N. magellanica* [@Nicastro2020-ih], in which detailed descriptions of the stable isotope analysis methodology can be found. In short, all shells were sectioned along the main growth axis to expose the internal growth structure. From the exposed section carbonate powder samples were acquired using a micromill with resolutions of 100–200 µm and by targeting the calcitic M+2 layer of the shells. Analytical precision of the isotope values was consistently at 0.05–0.10‰ and values are reported in per mil units (‰) relative to the VPDB (Vienna Pee Dee Belemnite) standard.

@fig-Nac_iso and @fig-Pat_iso show the previously acquired sequential δ^18^O-values of limpet shells by species. Due to the high sampling resolution, all shells show seasonally resolved and multi-year quasi-sinusoidal records of SST change , with *Nacella* sp. shells living 2–3 years and *Patella v*. regularly over 3 years. In those instances where longer lifetimes are encountered (i.e. specimen QG1-7246-1, @fig-Pat_iso b). Interestingly, the annual ranges in δ^18^O-values are not similar between the coeval *Nacella* sp. specimens with δ^18^O values for the previous summer temperatures ranging from 0.4‰ to above 1.0‰, and for winter values between 2.7‰ and 2.0‰. *P. vulgata* has similarly variable δ^18^O-values for summer and winter seasons, alas the lack of more that one modern specimens, means that we cannot quantify the variability of recorded annual ranges between specimens.

```{r Nacella_Isotope_plot}
#| fig-cap: Isotope values of *Nacella* sp. limpet shells from Cambaceres Bay (*a-f*).
#| fig-height: 10
#| fig-width: 10
#| label: fig-Nac_iso
# Filter data for Modern period
Nacella_iso <- D18O %>% filter(grepl("Nacella", species))

@@

# Loop over each sampleID to create individual plots
for (i in seq_along(sample_ids)) {
  sample_id <- sample_ids[i]
  plot <- Nacella_iso %>%
    filter(sampleID == sample_id) %>%
    ggplot(aes(dist, d18O)) +
    geom_path() +
    geom_point() +    labs(
      x = "Distance from margin in mm",
      y = expression(delta^18*O~("\u2030")),
      subtitle = sample_id,
      col = "",
      fill = ""
    ) +
    theme(plot.subtitle = element_text(hjust = 0.5))

@@

# Combine plots using patchwork and add letters
wrap_plots(plots,nrow=3) + plot_annotation(tag_levels = 'a')

@@

```{r Patella_Isotope_plot}
#| fig-cap: Isotope values of *Patella vulgata* limpet shells from Quoygrew (QG, *a-g*) and Rack Wick Bay (ORK, *h*) on Westray, Orkney.
#| fig-height: 10
#| fig-width: 10
#| label: fig-Pat_iso
# Filter data for Modern period
Patella_iso <- D18O %>% filter(grepl("Patella", species))

@@

@@

Because the exact location used in the analysis of stable oxygen isotopes is not possible to reconstruct for each carbonate powder sample, we used dynamic time warping (DTW) to align the time series of δ^18^O-values and Mg/Ca ratios. DTW is an algorithm that measures similarity between two proxy-sequences, which may vary in sampling resolution or interval. By stretching or compressing sections of the series, DTW finds the probable alignment between the two sequences. This allows us to compare the proxy data sets more effectively, ensuring that the temporal dynamics of each shell are accurately matched despite possible discrepancies in sampling intervals or rates. We applied the DTW algorithm using the `dtw` package in R [@Giorgino2009-sj; @R_Core_Team2020-mk], which provides a robust framework for aligning time series data. This involved selecting appropriate distance measures and constraints to ensure meaningful alignment. The process involved iterative adjustments to minimise the overall distance between corresponding points in the two data sets. The code used for this analysis as well as the data required to run it can be found in this **OSF repository**.

@@

```{r Patella_LIBS}
#| fig-cap: Example LIBS maps and line scans of analysed specimens. (*a+b*) ORK-LT and (*c*+*d*) QG1-7188-1.  Note that the ranges of Mg/Ca intensity ratios are not identical but are chosen arbitrarily to increase contrast. Areas in grey consist of low-Mg aragonite and fall outside the chosen colour-range.
#| fig-height: 10
#| fig-width: 15
#| label: fig-Pat_LIBS

@@

ORK_LT5_plot <- 
  ORK_LT5_map %>%
  filter(std < 0.15) %>%
  filter(mg_ca < 0.7 & mg_ca > 0.3) %>%
  ggplot() +
  aes(x, y, fill = mg_ca) +
  geom_tile(data = ORK_LT5_map, fill = "grey") +
  geom_tile(show.legend=FALSE) +
  geom_path(data=MgCa %>% filter(sampleID=="ORK_LT5"),aes(x,y),col="white",linetype="longdash")+
  theme_cowplot(20)+
  scale_fill_viridis_c(limits = c(0.3, 0.7),
                       breaks = seq(0.3, 0.7, by = 0.1)) +    
  theme(
    legend.position = c(0.45, 0.4),
    legend.title = element_text(hjust = 0.5)
    )+ # Center the legend title) +
    coord_fixed() +
    labs(x = "mm", y = "mm", fill = "Mg/Ca ")

ORK_LT5_line <- 
  MgCa %>% 
  filter(sampleID=="ORK_LT5") %>% 
  filter(relstd < 0.1) %>%
  filter(mg_ca < 0.6 & mg_ca > 0.3) %>%
  ggplot()+
  aes(dist,mg_ca,col=mg_ca)+
  geom_path(show.legend = FALSE)+
  scale_colour_viridis_c(limits = c(0.3, 0.6),
                       breaks = seq(0.3, 0.6, by = 0.1)) +    
  theme_cowplot(20)+
  labs(x = "Distance to edge [mm]", y = "Mg/Ca intensity ratio")

@@

QG1_7188_1_plot <- 
  QG1_7188_1_map %>%
  filter(relstd < 0.1) %>%
  filter(mg_ca < 0.6 & mg_ca > 0.3) %>%
  ggplot() +
  aes(x, y, col = mg_ca) +
  geom_point(data = QG1_7188_1_map, col = "grey") +
  geom_point(size=1,show.legend=FALSE) +
  geom_path(data=MgCa %>% filter(sampleID=="QG1-7188-1"),aes(x,y),col="white",linetype="longdash")+
  scale_colour_viridis_c(limits = c(0.3, 0.6),
                       breaks = seq(0.3, 0.6, by = 0.1)) +    
  theme_cowplot(20)+
  theme(
    legend.position = c(0.05, 0.78),
   legend.title = element_text(hjust = 0.5)
    )+ # Center the legend title) +
    coord_fixed() +
    labs(x = "mm", y = "mm", colour = "Mg/Ca")

QG1_7188_1_line <- 
  MgCa %>% 
  filter(sampleID=="QG1-7188-1") %>% 
  filter(relstd < 0.1) %>%
  filter(mg_ca < 0.6 & mg_ca > 0.3) %>%
  ggplot()+
  aes(dist,mg_ca,col=mg_ca)+
  geom_path(show.legend = FALSE)+
  scale_colour_viridis_c(limits = c(0.3, 0.6),
                       breaks = seq(0.3, 0.6, by = 0.1)) +    
  theme_cowplot(20)+
  labs(x = "Distance to edge [mm]", y = "Mg/Ca intensity ratio")

@@

Line scans also indicate well the quasi-sinusoidal change of Mg/Ca intensity ratios expected based on the stable isotope data and elemental-imaging. That said, some of the variability of Mg/Ca intensity ratios within one season such as the summer period of QG1-7188-1 between 0.5 and 2.0 mm distance to the shell edge (\@fig-Pat_LIBS d), is visible in a more dramatic manner than it appears in the 2D-image, showing the downside of line-scanning as opposed to the broader milling approach used for the analysis of carbonate powder [@Ferguson2011-zl].

@fig-Pat_Comp shows the line scan of ORK-LT5 in comparison to its δ^18^O-values (similar graphs for other specimen can be found in the supplementaries). The measurements of the distance to the shell edge are not entirely identical, most likely because one growth increment can have several different distances. Depending on where one measures the distance and a line scan towards the interior would be shorter than a line scan towards the exterior. That said, increases and decreases of both proxies seem to mirror each other and to be well aligned.

```{r Patella_Compare}
#| fig-cap: Comparison of LIBS line-scan and sequential δ^18^O-values of specimen ORK-LT5 
#| fig-height: 10
#| fig-width: 20
#| label: fig-Pat_Comp

P1 <-
  MgCa %>% 
  filter(sampleID=="ORK_LT5") %>% 
  filter(relstd < 0.1) %>%
  filter(mg_ca < 0.6 & mg_ca > 0.3) %>%
ggplot()+
  aes(dist,mg_ca)+
  geom_path()+
  theme_cowplot(20)+
  labs(x = "", y = "Mg/Ca intensity ratio")

P2 <-
  D18O %>% 
  filter(sampleID=="ORK_LT5") %>% 
  ggplot()+
  aes(dist,d18O)+
  geom_path()+
  theme_cowplot(20)+
  scale_y_reverse()+
  labs(x = "Distance to edge [mm]", y = expression("‰ δ" ^ "18" * "O [VPDB]"))

@@

Aligning the records programmatically using dynamic time warping, allowed us to compare them directly to determine specimen-specific equations and quantify the correlation of both proxies (@fig-Pat_Corr). The equations for all specimens are different but the majority (barring one: QG1-7246-1) do seem to cluster around shared parameters (δ^18^Ο = 0.6 – 0.065 \* Mg/Ca). The various coefficients of determination (R^2^) range between 0.81 (QG1-7246-1) and 0.95 (QG1-7189-2). These specimens are also the oldest and youngest, respectively, specimens, suggesting that lower growth rates and time averaging had a negative effect on the correlation of both proxies. The mean R^2^ value is 0.89, suggesting a generally good fit between the two proxies.

```{r Patella_Correlation_Graphs }
#| fig-cap: Correlation graphs for *Patella vulgata* specimens.
#| fig-height: 18
#| fig-width: 15
#| label: fig-Pat_Corr

>>>>

---
title: "Interspecies comparisons of Mg/Ca ratios in limpet shells"
author:
  - name: Niklas Hausmann
    affiliations:
      - name: Leibniz Zentrum für Archäologie
        address: Ludwig-Lindenschmit-Forum 1
        city: Mainz
        country: Germany
        postal-code: 55116
    orcid: 0000–0002–4381–8334
    email: niklas@palaeo.eu
    attributes:
      corresponding: true
  - name: Donna Surge
    affiliations:
      - name: University of North Carolina
        address: 104 South Road, 225 Geology Building
        city: Chapel Hill, NC
        country: US
        postal-code: 27599-3315
  - name: Ivan Briz-Godino
    affiliations:
      - name: 
        address: 
        city: 
        country: Spain
        postal-code: 
abstract: |
  This study provides a short reassessment of the use of Magnesium to Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface temperatures. While *Patella vulgata* along the Spanish shoreline has since then repeatedly produced reliable correlations between sea surface temperature and Mg/Ca ratios, this is not the case for other patelloid species. *Patella vulgata* and *Nacella deaureata* have been studied using Mg/Ca with mixed or contrary results. In this study, we present elemental maps of various such species together with stable oxygen isotope values. Our dataset also includes specimens that were previously unsuccessful in providing significant correlations between $\delta$^18^O and Mg/Ca ratios. By reassessing these previous specimens and including a wider range of modern and archaeological samples (n=12) from three patelloid species (*P. vulgata*, *N. deaureata*, and *N. magellanica*) we further add to the growing set of evidence for the reliable use of Mg/Ca ratios to detect palaeotemperature change and serve as a means to determine ontogenetic age and season of capture as well as to reveal locations of interest within the growth record (i.e. annual temperature minima and maxima) for the targeted analysis using $\delta$^18^O or clumped oxygen isotope analysis.
keywords: 
  - Sclerochronology
  - Limpets
  - Elemental Ratio
  - Mg/Ca
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    latex_engine: xelatex
    journal:
      name: Palaeo^3^
      formatting: preprint
      model: 3p
      cite-style: authoryear
output:
  units: cm

---

@@

  theme_set(theme_cowplot(10))

@@

Limpet shells are commonly found within archaeological sites and and past shorelines due to their robust carbonate structure and the long-term use of limpets as a marine food source. They have successfully been studied in the past in the context of coastal subsistence economies, site occupation on seasonal [@Shackleton1973-ij; @Parker2018-wf; @Bosch2018-ud] or long-term scale [@Ortiz2015-mr], as well as palaeotemperature [@Fenger2007-gf; @Surge2012-ba; @Wang2012-ee; @Colonese2012-ct; @Ferguson2011-zl]. Determining past sea surface temperature (SST) change largely relies on the measurement of $\delta$^18^O-values within the calcitic parts of the shell, but attempts have been made to also use elemental ratios, such as magnesium to calcium (Mg/Ca), to have an alternative measure, that potentially provides a more accurate SST estimate than $\delta$^18^O-values, which are also affected by changes in salinity. In addition, the data acquisition of elemental ratios — either through laser-ablation-isotope-ratio-mass-spectrometry (LA-ICP-MS) or laser induced breakdown spectroscopy (LIBS), can be much faster and cost-effective, increasing the number of specimens that can be studied overall [@Durham2017-fh; @Hausmann2023-ih].

Problematically, links between Mg/Ca ratios and SST changes in many marine mollusc shells, including limpets, have shown to be unreliable [@Surge2008-ri; @Wanamaker2008-zl; @Schone2010-yl; @Freitas2012-tx; @Graniero2017-io; @Poulain2015-dg; @Vihtakari2017-wd]. This is particularly the case where there is little available additional information on metabolic processes, organic components of the shell matrix, intra-increment and intra-shell variability, and growth rates, which can independently and unpredictably affect Mg/Ca ratios and confound their interpretation as temperature proxy. Confusingly, multiple different temperature equations have been found for the same species (see [@Freitas2012-tx; @Vihtakari2017-wd] and references therein). Coeval specimens sharing one locality can also show differences in their relation to SST [@Hausmann2019-fi]. Where the use of Mg/Ca as a palaeotemperature proxy was successful, anomalous patterns in some specimens had still to be filtered out by hand, reducing the overall robustness of the results of those successful studies [@Ferguson2011-zl].

@@

By relying and adding onto published datasets, we were able to simultaneously avoid costs for new high-resolution $\delta$^18^O-data, to use real-world examples, and to also provide pilot-data for areas of existing research interest. Generally establishing the usefulness and reliability of Mg/Ca as SST proxy in limpets should help to provide a platform for future research and an important stepping stone to better understand elemental ratios in other marine mollusc shells.

@@

::: {#Table_1}
\fontsize{8pt}{8pt}\selectfont

| Context        | Study              | Species          | Location        | Sample ID  | Previous analyses |
|------|------------|-------|--------|------|----------|
| Modern         | [@Nicastro2020-ih] | *N. deaureata*   | Cambaceres Bay  | ND-1016-3  | Stable oxygen     |
|                |                    |                  | \- Tierra del   | ND-1016-4  | and carbon        |
|                |                    | *N. magellanica* | Fuego (AR)      | NM-1016-1  | isotope analysis  |
|                |                    |                  |                 | NM-1016-3  |                   |
|                |                    |                  |                 |            |                   |
|                | [@Graniero2017-io] | *P. vulgata*     | Rack Wick Bay - | ORK-LT5    | Stable oxygen     |
|                |                    |                  | Westray (UK)    |            | and carbon        |
|                |                    |                  |                 |            | isotope analysis; |
|                |                    |                  |                 |            | Mg, Li, Sr, Ca.   |
|                |                    |                  |                 |            |                   |
| Archaeological | [@Surge2012-ba]    | *P. vulgata*     | Rack Wick Bay - | QG1-7188-1 | Stable oxygen     |
|                |                    |                  | Westray (UK)    | QG1-7189-2 | and carbon        |
|                |                    |                  |                 | QG2-1061-1 | isotope analysis  |
|                |                    |                  |                 | QG2-1064-1 |                   |
|                |                    |                  |                 | QG2-7180-1 |                   |
|                |                    |                  |                 | QG2-7180-2 |                   |
|                |                    |                  |                 | QG1-7246-1 |                   |

\normalsize
There are two main locations that we sourced specimens from: the Beagle Channel in Tierra del Fuego (Argentina) and the island of West Ray in Orkney (UK). Four specimens were collected from the Beagle Channel in Cambaceres Bay in October 2016. The area receives around 570 mm precipitation but the channel waters are predominantly influenced by marine currents and in the period of October 2015 to October 2016 stayed around 30.7 ± 0.7 psu [@Nicastro2020-ih]. SSTs range from `r CamBayMin`ºC in August to `r CamBayMax`ºC in February.

@@

```{r SST_plot,fig.pos='H'}
#| fig-cap: Monthly Sea Surface Temperature values averaged from  years. SST values were sourced from @Good2020-nl.
#| fig-height: 3
#| fig-width: 6
#| label: fig-SSTs

# Plot the data
SST_df %>% 
    group_by(dataset, month) %>%
  summarise(sd_sst = sd(mean_sst, na.rm = TRUE),
            mean_sst = mean(mean_sst, na.rm = TRUE)) %>%
ggplot(aes(x = month, y = mean_sst, colour = dataset,group=dataset)) +
  geom_ribbon(aes(ymin = mean_sst - sd_sst, ymax = mean_sst + sd_sst,fill=dataset), alpha = 0.2,col="white") +
  geom_path() +
  scale_colour_manual(values = c( "#007f7f", "#EC7063")) +
scale_fill_manual(values = c( "#007f7f", "#EC7063"))+
  labs(title = "",
       x = "Month",
       y = "Mean Analysed SST (°C)",colour="",fill="") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1),
          legend.position = c(0.1, 0.9))+  
  scale_y_continuous(breaks = c(5, 10, 15),limits = c(3,15))

@@

```{r Nacella_Isotope_plot,dev='png',dpi=300}
#| fig-cap: Isotope values of *Nacella* sp. limpet shells from Cambaceres Bay.
#| fig-height: 5
#| fig-width: 10
#| label: fig-Nac_iso
# Filter data for Modern period
Nacella_iso <- D18O %>% filter(grepl("Nacella", species))

@@

# Loop over each sampleID to create individual plots
for (i in seq_along(sample_ids)) {
  sample_id <- sample_ids[i]
  plot <- Nacella_iso %>%
    filter(sampleID == sample_id) %>%
    ggplot(aes(dist, d18O)) +
    geom_path() +
    geom_point() +    labs(
      x = "Distance from margin in mm",
      y = expression(paste(delta^{18},"O (\u2030)")),
      subtitle = sample_id,
      col = "",
      fill = ""
    ) +
    theme(plot.subtitle = element_text(hjust = 0.5))

@@

# Combine plots using patchwork and add letters
wrap_plots(plots,nrow=2) + plot_annotation(tag_levels = 'a')

@@

High-quality oxygen isotope values ($\delta$^18^O) were acquired from previous publications on *P. vulgata* [@Surge2012-ba; @Graniero2017-io], *N. deaureata* and *N. magellanica* [@Nicastro2020-ih], in which detailed descriptions of the stable isotope analysis methodology can be found. In short, all shells were sectioned along the main growth axis to expose the internal growth structure. From the exposed section carbonate powder samples were acquired using a micromill with resolutions of 100–200 µm and by targeting the calcitic M+2 layer of the shells. Analytical precision of the isotope values was consistently at 0.05–0.10‰ and values are reported in per mil units (‰) relative to the VPDB (Vienna Pee Dee Belemnite) standard.

@fig-Nac_iso and @fig-Pat_iso show the previously acquired sequential $\delta$^18^O-values of limpet shells by species. Due to the high sampling resolution, all shells show seasonally resolved and multi-year quasi-sinusoidal records of SST change , with *Nacella* sp. shells living 2–3 years and *Patella v*. regularly over 3 years. In those instances where longer lifetimes are encountered (i.e. specimen QG1-7246-1, @fig-Pat_iso b).

```{r Patella_Isotope_plot,dev='png',dpi=300,fig.pos='H'}
#| fig-cap: Isotope values of *Patella vulgata* limpet shells from Quoygrew (QG, *a-g*) and Rack Wick Bay (ORK, *h*) on Westray, Orkney.
#| fig-height: 10
#| fig-width: 10
#| label: fig-Pat_iso
# Filter data for Modern period
Patella_iso <- D18O %>% filter(grepl("Patella", species))

@@

 Interestingly, the annual ranges in $\delta$^18^O-values are not similar between the coeval *Nacella* sp. specimens with $\delta$^18^O values for the previous summer temperatures ranging from 0.4‰ to above 1.0‰, and for winter values between 2.7‰ and 2.0‰. *P. vulgata* has similarly variable $\delta$^18^O-values for summer and winter seasons, alas the lack of more that one modern specimens, means that we cannot quantify the variability of recorded annual ranges between specimens.

@@

Because the exact location used in the analysis of stable oxygen isotopes is not possible to reconstruct for each carbonate powder sample, we used dynamic time warping (DTW) to align the time series of $\delta$^18^O-values and Mg/Ca ratios. DTW is an algorithm that measures similarity between two proxy-sequences, which may vary in sampling resolution or interval. By stretching or compressing sections of the series, DTW finds the probable alignment between the two sequences. This allows us to compare the proxy data sets more effectively, ensuring that the temporal dynamics of each shell are accurately matched despite possible discrepancies in sampling intervals or rates. We applied the DTW algorithm using the `dtw` package in R [@Giorgino2009-sj; @R_Core_Team2020-mk], which provides a robust framework for aligning time series data. This involved selecting appropriate distance measures and constraints to ensure meaningful alignment. The process involved iterative adjustments to minimise the overall distance between corresponding points in the two data sets.

@@

```{r Patella_LIBS,fig.pos='H'}
#| fig-cap: Example LIBS maps and line scans of analysed specimens. (*a+b*) ORK-LT and (*c*+*d*) QG1-7188-1.  Note that the ranges of Mg/Ca intensity ratios are not identical but are chosen arbitrarily to increase contrast. Areas in grey consist of low-Mg aragonite and fall outside the chosen colour-range.
#| fig-height: 5
#| fig-width: 10
#| label: fig-Pat_LIBS

@@

ORK_LT5_plot <- 
  ORK_LT5_map %>%
  filter(std < 0.15) %>%
  filter(mg_ca < 0.7 & mg_ca > 0.3) %>%
  ggplot() +
  aes(x, y, fill = mg_ca) +
  geom_tile(data = ORK_LT5_map, fill = "grey") +
  geom_tile(show.legend=FALSE) +
  geom_path(data=MgCa %>% filter(sampleID=="ORK_LT5"),aes(x,y),col="white",linetype="longdash")+
  theme_cowplot(8)+
  scale_fill_viridis_c(limits = c(0.3, 0.7),
                       breaks = seq(0.3, 0.7, by = 0.1)) +    
  theme(
    legend.position = c(0.45, 0.4),
    legend.title = element_text(hjust = 0.5)
    )+ # Center the legend title) +
    coord_fixed() +
    labs(x = "mm", y = "mm", fill = "Mg/Ca ")

ORK_LT5_line <- 
  MgCa %>% 
  filter(sampleID=="ORK_LT5") %>% 
  filter(relstd < 0.1) %>%
  filter(mg_ca < 0.6 & mg_ca > 0.3) %>%
  ggplot()+
  aes(dist,mg_ca,col=mg_ca)+
  geom_path(show.legend = FALSE)+
  scale_colour_viridis_c(limits = c(0.3, 0.6),
                       breaks = seq(0.3, 0.6, by = 0.1)) +    
  theme_cowplot(8)+
  labs(x = "Distance to edge [mm]", y = "Mg/Ca intensity ratio")

@@

QG1_7188_1_plot <- 
  QG1_7188_1_map %>%
  filter(relstd < 0.1) %>%
  filter(mg_ca < 0.6 & mg_ca > 0.3) %>%
  ggplot() +
  aes(x, y, col = mg_ca) +
  geom_point(data = QG1_7188_1_map, col = "grey",size=0.2) +
  geom_point(size=0.2,show.legend=FALSE) +
  geom_path(data=MgCa %>% filter(sampleID=="QG1-7188-1"),aes(x,y),col="white",linetype="longdash")+
  scale_colour_viridis_c(limits = c(0.3, 0.6),
                       breaks = seq(0.3, 0.6, by = 0.1)) +    
  theme_cowplot(8)+
  theme(
    legend.position = c(0.05, 0.78),
   legend.title = element_text(hjust = 0.5)
    )+ # Center the legend title) +
    coord_fixed() +
    labs(x = "mm", y = "mm", colour = "Mg/Ca")

QG1_7188_1_line <- 
  MgCa %>% 
  filter(sampleID=="QG1-7188-1") %>% 
  filter(relstd < 0.1) %>%
  filter(mg_ca < 0.6 & mg_ca > 0.3) %>%
  ggplot()+
  aes(dist,mg_ca,col=mg_ca)+
  geom_path(show.legend = FALSE)+
  scale_colour_viridis_c(limits = c(0.3, 0.6),
                       breaks = seq(0.3, 0.6, by = 0.1)) +    
  theme_cowplot(8)+
  labs(x = "Distance to edge [mm]", y = "Mg/Ca intensity ratio")

@@

Line scans also indicate well the quasi-sinusoidal change of Mg/Ca intensity ratios expected based on the stable isotope data and elemental-imaging. That said, some of the variability of Mg/Ca intensity ratios within one season such as the summer period of QG1-7188-1 between 0.5 and 2.0 mm distance to the shell edge (@fig-Pat_LIBS d), is visible in a more dramatic manner than it appears in the 2D-image, showing the downside of line-scanning as opposed to the broader milling approach used for the analysis of carbonate powder [@Ferguson2011-zl].

@fig-Pat_Comp shows the line scan of ORK-LT5 in comparison to its $\delta$^18^O-values (similar graphs for other specimen can be found in the supplementaries). The measurements of the distance to the shell edge are not entirely identical, most likely because one growth increment can have several different distances. Depending on where one measures the distance and a line scan towards the interior would be shorter than a line scan towards the exterior. That said, increases and decreases of both proxies seem to mirror each other and to be well aligned.

```{r Patella_Compare,dev='png',dpi=300}
#| fig-cap: Comparison of LIBS line-scan and sequential $\delta$^18^O-values of specimen ORK-LT5 
#| fig-height: 7
#| fig-width: 16
#| label: fig-Pat_Comp

P1 <-
  MgCa %>% 
  filter(sampleID=="ORK_LT5") %>% 
  filter(relstd < 0.1) %>%
  filter(mg_ca < 0.6 & mg_ca > 0.3) %>%
ggplot()+
  aes(dist,mg_ca)+
  geom_path()+
  theme_cowplot(17)+
  labs(x = "", y = "Mg/Ca intensity ratio")

P2 <-
  D18O %>% 
  filter(sampleID=="ORK_LT5") %>% 
  ggplot()+
  aes(dist,d18O)+
  geom_path()+
  theme_cowplot(18)+
  scale_y_reverse()+
  labs(x = "Distance to edge [mm]", y = expression(paste(delta^{18},"O (\u2030)")))

@@

Aligning the records programmatically using dynamic time warping, allowed us to compare them directly to determine specimen-specific equations and quantify the correlation of both proxies (@fig-Pat_Corr). The equations for all specimens are different but the majority (barring one: QG1-7246-1) do seem to cluster around shared parameters ($\delta$^18^Ο = 0.6 – 0.065 \* Mg/Ca). The various coefficients of determination (R^2^) range between 0.81 (QG1-7246-1) and 0.95 (QG1-7189-2). These specimens are also the oldest and youngest, respectively, specimens, suggesting that lower growth rates and time averaging had a negative effect on the correlation of both proxies. The mean R^2^ value is 0.89, suggesting a generally good fit between the two proxies.

```{r Patella_Correlation_Graphs,dev='png',dpi=300}
#| fig-cap: Correlation graphs for *Patella vulgata* specimens.
#| fig-height: 8
#| fig-width: 16
#| label: fig-Pat_Corr

P_all <- Corr_data %>% 
  pivot_wider(values_from = "value",names_from = "proxy") %>% 
  # mutate(region = substr(sampleID, 1, 1)) %>% 
  left_join(D18O %>% select(sampleID,period,species) %>% unique(),by="sampleID") %>% 
  filter(species=="Patella vulgata") %>% 
  ggplot()+
  aes(d18o,mg_ca,col=sampleID)+
  geom_point(show.legend = FALSE)+
  theme_cowplot(20)+
  labs(x=expression(paste(delta^{18},"O (\u2030)")),y="Mg/Ca intensity ratio")+
  stat_smooth(aes(fill = sampleID,col=sampleID), method = "lm", formula = y~x,show.legend = FALSE) +
  # stat_regline_equation(label.x = 2, #label.y = 32
  #   aes(label =  paste(after_stat(eq.label), after_stat(adj.rr.label), sep = "~~~~")),
  #   formula = y~x
  # ) +
  scale_colour_manual(values = c("#542788","#007f7f", "#77b0a3","black", "#1f78b4","grey50", "#d73027", "#EC7063")) +
scale_fill_manual(values = c( "#542788","#007f7f", "#77b0a3","black", "#1f78b4","grey50", "#d73027", "#EC7063"))#+
  # facet_wrap(~region,ncol=2,scales = "free")
